---
title: Analysis of BMI and All-Cause Morality in NHANES III and NHANES 1999-2004
---

We first load several required `R` packages and pre-parsed data.  Data come from NHANES III and continuous NHANES (cycles 1-3, 1999-2004), and followup mortality data from 2006 and 2011.  The model below uses only 2006 mortality data.  

```{r setup}
library(tidyverse)
library(cowplot)
library(broom)
library(survey)
library(survival)
library(stringr)

load('2016-10-06.RData')
```

We will examine only individuals at least 50 years old and less than 70 years old.  We censor mortality data for individuals who are 70 years old or more.   

```{r}
dataf = dataf %>% 
	## Drop entries with NA id or 0 sample weight
	filter(!is.na(id), sample.weight > 0) %>%
	mutate(
		## Censor mortality at the 85-years-old cutoff: 
		##  actual age at followup
		age.follow.2006 = age.months + follow.months.2006,
		age.follow.2011 = age.months + follow.months.2011,
		##  censor at 85 years old
		age.follow.2006.c = pmin(age.follow.2006, 70*12+11),
		age.follow.2011.c = pmin(age.follow.2011, 70*12+11),
		##  censored followup months
		follow.months.2006.c = age.follow.2006.c - age.months,
		follow.months.2011.c = age.follow.2011.c - age.months,
		## censored mortality
		mort.2006.c = ifelse(age.follow.2006 <= 70*12+11, mort.2006, NA), 
		mort.2011.c = ifelse(age.follow.2011 <= 70*12+11, mort.2011, NA),
		## For cross-referencing, number the rows in the master dataset
		row.num = row_number())
```

We filter based on the following criteria:  

- Smoker status (nonsmokers only)
- Age (at least 50 years old, less than 70 years old)
- Followup time greater than 0 months
- Data available on BMI status, education, and mortality

After filtering, we move to a complex survey design object.

```{r}
dataf.2006 = subset(dataf, (!smoker) & (age.months >= 50*12 + 12) & 
						(age.months < 70*12 + 12) & 
						(follow.months.2006 > 0) &
						# (bmi.cat != 'underweight') & 
						(bmi < 75) & (bmi.max < 75) &
						!is.na(bmi) &
						!is.na(education) & !is.na(mort.2006))

## Move to a survey design object
design_unfltd = svydesign(id = ~ psu, strata = ~ stratum, weights = ~ sample.weight, 
						  nest = TRUE, 
						  data = dataf)
design.2006 = subset(design_unfltd, row.num %in% dataf.2006$row.num)
```

The output below gives a descriptive summary of the dataset used to fit the model. 

```{r}
summary(dataf.2006)
```

Next we fit a Cox proportional hazards model, using categorical BMI.  

```{r}
coxfit.cat = svycoxph(Surv(age.years, mort.2006.c) ~ 
				  			sex + race.ethnicity + education +
				  			bmi.cat, 
						   design = design.2006)
summary(coxfit.cat)
```

To graphically check model fit, we compare Kaplan-Meier estimates derived from the data themselves (black) and the fitted Cox PH model (red; 95% confidence interval indicated by red shading).  There appears to be somewhat good agreement between the curve estimates, although there is almost total uncertainty about the Cox PH model's estimate at the high end of the age range.  

```{r}
km.obs = svykm(Surv(age.years, mort.2006.c) ~ 1, 
			   design = design.2006, se = FALSE)
km.model = survfit(coxfit.cat, se.fit = TRUE)
kms = data.frame(
	age = km.obs$time[-1],
	obs = km.obs$surv[-1],
	model.pe = km.model$surv,
	model.upper = km.model$upper, 
	model.lower = km.model$lower)
kms.plot = ggplot(kms) + 
	geom_line(aes(age, obs), color = 'black') +
	geom_line(aes(age, model.pe), color = 'red') +
	geom_ribbon(aes(age, ymin = model.lower, ymax = model.upper), fill = 'red', 
				alpha = .25)
kms.plot
```

We also plot deviance residuals against (a) the model's predictions, to check for outliers and heteroscedasticity; and (b) the model's age, to check the proportional hazards assumption.  LOESS curves are used to check independence; these curves should be flat.  There are no obvious outliers, and the residuals appear to be homoscedastic; there is a slight bend in the LOESS curve above about 62 years old.  

```{r}
res.df = data.frame(
	age = coxfit.cat$model$`Surv(age.years, mort.2006.c)`[,1],
	prediction = coxfit.cat$linear.predictors,
	residual = residuals(coxfit.cat, type = 'deviance')
	)
ggplot(res.df, aes(prediction, residual)) + geom_point(position = 'jitter') +
	geom_smooth()
ggplot(res.df, aes(age, residual)) + geom_point() + geom_smooth()
```

Next we extract model coefficients.  

```{r}
coeffs = tidy(coxfit.cat) %>%
	transmute(term = term, 
			  estimate = exp(estimate),
			  conf.low = exp(conf.low),
			  conf.high = exp(conf.high)) %>%
	mutate(term = str_replace(term, 'sex', ''),
		   term = str_replace(term, 'race.ethnicity', ''),
		   term = str_replace(term, 'education', ''),
		   term = str_replace(term, 'bmi.cat1', 'underweight'),
		   term = str_replace(term, 'bmi.cat3', 'overweight'),
		   term = str_replace(term, 'bmi.cat4', 'obese I'), 
		   term = str_replace(term, 'bmi.cat5', 'obese II')) %>%
	mutate(term = factor(term, levels = term, ordered = TRUE), 
		   group = c('sex', 
		   		  rep('race.ethnicity', 3), 
		   		  rep('education', 4), 
		   		  rep('bmi', 4)))

ggplot(coeffs, aes(term, estimate, color = group)) + 
	geom_hline(yintercept = 1) +
	geom_point() + 
	geom_linerange(aes(ymin = conf.low, ymax = conf.high)) +
	coord_flip()
```
			  
			  
			  